Impacts of Isolation by Distance on Demographic Modeling and Association Studies

### 2019-03-15 ms/spaceness.tex
new gwas fig w pca var

%Simulating Isolation by Distance in Continuous Space
%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
\title{Space is the Place: Impacts of Continuous Spatial Structure on Demographic Modeling and Association Studies}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20. In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis. We demonstrate that stratification bias in continuous populations is worst when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: example phenotype distributions and sampling density. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. Note the log scale. D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1). The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect inflated -log10(p).}
\label{fig:spectrum}
\end{figure*}

>>>>

%Simulating Isolation by Distance in Continuous Space
%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
\title{Impacts of Isolation by Distance on Demographic Modeling and Association Studies}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum is biased towards inferring a turbulent past demographic history and a recent decline in populations when neighborhood sizes are below 20. In association analyses spatially correlated environments cause genome-wide signals of genetic association with purely environmental phenotypes, and this bias is only partially corrected by regressing out principal components positions during the analysis. We find that association analyses in continuous populations are most biased when environmental effects are spatially discrete, but that corrections for clinal phenotypes can also fail under high dispersal because PCA describes less total variation in the data when isolation by distance is weak.

@@

\afterpage{\clearpage}
\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots_logX.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on linear regression GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: example phenotype distributions and sampling density. B: proportion of total variance explained by the first 10 PC axes, by neighborhood size. C: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. D: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution for neighborhood sizes under 65 ($\sigma$=1). The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect inflated -log10(p).}
\label{fig:spectrum}
\end{figure*}
